---
title: "Data visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = TRUE,
  echo = TRUE,
  comment = "#>",
  dpi = 150,
  fig.align = "center",
  out.width = "80%",
  fig.height = 4,
  fig.width = 7
)
```

The package `forcis` provides [numerous functions](https://frbcesab.github.io/forcis/reference/index.html#visualization-tools) to visualize FORCIS data. This vignette shows how to use and customize these functions.


## Setup

First, let's import the required packages.

```{r setup}
library(forcis)
library(ggplot2)
```

Before proceeding, let's download the latest version of the FORCIS database.

```{r 'download-db', eval=FALSE}
# Create a data/ folder ----
dir.create("data")

# Download latest version of the database ----
download_forcis_db(path = "data", version = NULL)
```

The vignette will use the plankton net data of the FORCIS database. Let's import the latest release of the data.

```{r 'load-data', echo=FALSE}
file_name <- system.file(
  file.path("extdata", "FORCIS_net_sample.csv"),
  package = "forcis"
)
net_data <- read.csv(file_name)
```

```{r 'load-data-user', eval=FALSE}
# Import net data ----
net_data <- read_plankton_nets_data(path = "data")
```

**NB:** In this vignette, we use a subset of the plankton net data, not the whole dataset.


## Spatial visualization

### Creating a world map

The function `geom_basemap()` can be used to easily add World countries, oceans and bounding box to a `ggplot2` object.

```{r 'geom-basemap'}
# World basemap ----
ggplot() +
  geom_basemap()
```

These layers come from the [Natural Earth](https://www.naturalearthdata.com/) website and are defined in the [Robinson projection](https://epsg.io/54030).



### Mapping FORCIS data

The function `ggmap_data()` can be used to plot FORCIS data on a World map. Let's map the plankton nets data.

```{r 'ggmap-raw-data'}
# Map raw net data ----
ggmap_data(net_data)
```

User can customize the aesthetic of the data:

```{r 'ggmap-raw-data-2'}
# Customize map ----
ggmap_data(net_data, col = "#ff0000", fill = NA, shape = 21, size = 3)
```

This function works with the output of various functions available in the `forcis` package. For example:

```{r 'ggmap-filtered-data', echo=TRUE, eval=FALSE}
# Filter net data ----
indian_net_data <- filter_by_ocean(net_data, ocean = "Indian Ocean")

# Map filtered data ----
ggmap_data(indian_net_data)
```

Note that the `forcis` package is pipe-friendly.

```{r 'ggmap-filtered-data-pipe', eval=FALSE}
# Same as before, but w/ the pipe ----
net_data %>%
  filter_by_ocean(ocean = "Indian Ocean") %>%
  ggmap_data()
```

You can export this map with the function `ggsave()` of the package `ggplot2`.


```{r 'ggmap-save', eval=FALSE}
# Map filtered data ----
indian_net_data_map <- net_data %>%
  filter_by_ocean(ocean = "Indian Ocean") %>%
  ggmap_data() +
  ggtitle("FORCIS net data - Indian Ocean")

# Save as PNG ----
ggsave(
  indian_net_data_map,
  filename = "indian_net_data_map.png",
  width = 20,
  height = 11,
  units = "cm",
  dpi = 300,
  scale = 1.5,
  bg = "white"
)
```


## Temporal visualization

### Plot data by year of sampling

The function `plot_record_by_year()` plots the number of records (y-axis) by year (x-axis).


```{r 'plot-record-by-year'}
# Plot number of records by year ----
plot_record_by_year(net_data)
```



### Plot data by month of sampling

The function `plot_record_by_month()` plots the number of records (y-axis) by month (x-axis).

```{r 'plot-record-by-month'}
# Plot number of records by month ----
plot_record_by_month(net_data)
```



### Plot data by season

The function `plot_record_by_season()` plots the number of records (y-axis) by season (x-axis).


```{r 'plot-record-by-season'}
# Plot number of records by season ----
plot_record_by_season(net_data)
```


## Vertical visualization

### Plot data by depth of sampling

The function `plot_record_by_depth()` plots the number of records (x-axis) by depth intervals (y-axis).


```{r 'plot-record-by-depth'}
# Plot number of records by depth ----
plot_record_by_depth(net_data)
```
