---
title: "Database versions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Database versions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval     = FALSE,
  echo     = TRUE,
  comment  = "#>"
)
```

The [FORCIS database](https://zenodo.org/doi/10.5281/zenodo.7390791) is regularly updated. The global structure of the `csv` files doesn't change between versions but some bugs can be fixed and new records can also be added. This is why it is **recommended to use the latest version** of the database.


## Using the latest version

The package `forcis` is designed to handle the versioning of the database on Zenodo and will inform users if a new version is available each time they use one of the `read_*_data()` functions. The following message will be displayed:

```
A newer version of the FORCIS database is available. Use 'download_forcis_db(version = NULL)' to download it.
```

As mentioned in this message, users can choose whether to download the latest version of the database.

A **good practice** to ensure that the user works with the latest version of the database might be to add this line at the beginning of the script:

```{r 'use-latest-version'}
download_forcis_db(version = NULL, ...)
```

By setting `version = NULL`, the function `download_forcis_db()` will download the latest version of the FORCIS database from Zenodo (if not already downloaded). If no newer version has been released on Zenodo and if the latest version of the database has already been downloaded, this step will be ignored.

Note that the FORCIS database is saved in `forcis-db/version-99/`, where `99` is the label of the version. If you regularly download the new versions, the content of the folder `forcis-db/` may grow over time and it can be interesting to manually delete the folder of the previous versions.


## Using a specific version

If for some reason users want to freeze the version they use, it is possible to disable this message by setting the argument `check_for_update` to `FALSE` in the `read_*_data()` functions.

It is also possible to disable this message globally for the current session:

```{r 'disable-message'}
options(check_for_update = FALSE)
```

The package `forcis` _"knows"_ which version of the database you used last time. A hidden file named `.forcis` is created (or updated) each time the function `download_forcis_db()` or `read_*_data()` is called. This hidden file contains one line:

```
FORCIS_VERSION=99
```

Note that this hidden file is used by some functions of the `forcis` package. Users don't need to edit it manually.

If you want to be sure to use a specific version of the database, you can use the argument `version` of the `read_*_data()` functions, or use the following line:

```{r 'freeze-version'}
options(forcis_version = "99")
```

At any time, user can run the function `get_current_version()` to check the version of the database currently in use.

> But again, it is **recommended to use the latest version** of the database.
