---
title: "Data visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  eval      = TRUE,
  echo      = TRUE,
  comment   = "#>",
  dpi       = 150,
  fig.align = "center",
  out.width = "90%"
)
```

> **NOTE:** THIS IS A WORK IN PROGRESS.

The package `forcis` provides [a lot of functions](https://frbcesab.github.io/forcis/reference/index.html#visualization-tools) to visualize FORCIS data. This vignette shows how to use and customize these functions.


## Setup

First, let's import the required packages.

```{r setup}
library(forcis)
library(ggplot2)
```

Before going any further, we will download the latest version of the FORCIS database.

```{r 'download-db', eval=FALSE}
# Create a data/ folder ----
dir.create("data")

# Download latest version of the database ----
download_forcis_db(path = "data", version = NULL)
```

The vignette will use the PUMP data of the FORCIS database. Let's import the latest release of the data.

```{r 'load-data', echo=FALSE}
file_name <- system.file(file.path("extdata", "FORCIS_pump_sample.csv"), package = "forcis")
pump_data <- vroom::vroom(file_name, delim = ";", show_col_types = FALSE)
```

```{r 'load-data-user', eval=FALSE}
# Import pump data ----
pump_data <- read_pump_data(path = "data")
```

**NB:** In this vignette, we use a subset of the PUMP data, not the whole dataset.


## World base map

The function `geom_basemap()` can be used to easily add World countries, oceans and bounding box to a `ggplot2` object.

```{r 'geom-basemap', echo=TRUE, fig.height=11, fig.width=20}
# World basemap ----
ggplot() +
  geom_basemap()
```

These layers come from the [Natural Earth](https://www.naturalearthdata.com/) website and are defined in the [Robinson projection](https://epsg.io/54030).



## Map FORCIS data

The function `ggmap_data()` can be used to plot FORCIS data on a World map. Let's map the PUMP data.

```{r 'ggmap-raw-data', echo=TRUE, fig.height=11, fig.width=20}
# Map raw pump data ----
ggmap_data(pump_data)
```

User can customize the aesthetic of the data:

```{r 'ggmap-raw-data-2', echo=TRUE, fig.height=11, fig.width=20}
# Customize map ----
ggmap_data(pump_data, col = "black", fill = "red", shape = 21, size = 5)
```

And add a title:

```{r 'ggmap-raw-data-3', echo=TRUE, fig.height=11, fig.width=20}
# Add a title ----
ggmap_data(pump_data) +
  ggtitle("FORCIS Pump data")
```

This function works with the output of various functions available in the `forcis` package. For example:

```{r 'ggmap-filtered-data', echo=TRUE, fig.height=11, fig.width=20}
# Filter pump data ----
indian_pump_data <- filter_by_ocean(pump_data, ocean = "Indian Ocean")

# Map filtered data ----
ggmap_data(indian_pump_data)
```

Note that the `forcis` package is pipe-friendly.

```{r 'ggmap-filtered-data-pipe', eval=FALSE}
# Same as before, but w/ the pipe ----
pump_data %>%
  filter_by_ocean(ocean = "Indian Ocean") %>%
  ggmap_data()
```

You can export this map with the function `ggsave()` of the package `ggplot2`.


```{r 'ggmap-save', eval=FALSE}
# Map filtered data ----
indian_pump_data_map <- pump_data %>%
  filter_by_ocean(ocean = "Indian Ocean") %>%
  ggmap_data() +
  ggtitle("FORCIS Pump data - Indian Ocean")

# Save as PNG ----
ggsave(indian_pump_data_map, 
       filename = "indian_pump_data_map.png",
       width    = 20, 
       height   = 11, 
       units    = "cm", 
       dpi      = 300,
       scale    = 1.5,
       bg       = "white")
```
