---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval     = FALSE,
  echo     = TRUE,
  comment  = "#>"
)
```


> **NOTE:** THIS IS A WORK IN PROGRESS.

Add an good introduction of the package.


## Installation

To install the `forcis` package, run:

```{r 'installation', eval=FALSE}
## Install < remotes > package (if not already installed) ----
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

## Install dev version of < forcis > from GitHub ----
remotes::install_github("FRBCesab/forcis")
```

The `forcis` package depends on the [`sf`](https://r-spatial.github.io/sf/) package which requires some spatial system libraries (GDAL and PROJ). Please read [this page](https://github.com/r-spatial/sf?tab=readme-ov-file#installing) if you have any trouble to install `forcis`.



## Setup

First, let's attach the required packages.

```{r setup}
library(forcis)
```

## Download FORCIS database

Before going any further, we will download the latest version of the FORCIS database.

```{r 'download-db', eval=FALSE}
# Create a data/ folder ----
dir.create("data")

# Download latest version of the database ----
download_forcis_db(path = "data", version = NULL)
```


## Import FORCIS data

The vignette will use the plankton nets data of the FORCIS database. Let's import the latest release of the data.

```{r 'load-data', echo=FALSE}
file_name <- system.file(file.path("extdata", "FORCIS_net_sample.csv"), package = "forcis")
net_data <- read.table(file_name, dec = ".", sep = ";")
```

```{r 'load-data-user', eval=FALSE}
# Import plankton nets data ----
net_data <- read_plankton_nets_data(path = "data")
```

**NB:** In this vignette, we use a subset of the plankton nets data, not the whole dataset.


## Another section

[...]
