---
title: "Select, reshape, and filter data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Select, reshape, and filter data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval     = TRUE,
  echo     = TRUE,
  comment  = "#>",
  dpi       = 120,
  fig.align = "center",
  out.width = "80%"
)
```

The package `forcis` provides [a lot of functions](https://frbcesab.github.io/forcis/reference/index.html#select-and-filters-tools) to filter, reshape, and select FORCIS data. This vignette shows how to use these functions.


## Setup

First, let's import the required packages.

```{r setup}
library(forcis)
```

Before going any further, we will download the latest version of the FORCIS database.

```{r 'download-db', eval=FALSE}
# Create a data/ folder ----
dir.create("data")

# Download latest version of the database ----
download_forcis_db(path = "data", version = NULL)
```

The vignette will use the plankton nets data of the FORCIS database. Let's import the latest release of the data.

```{r 'load-data', echo=FALSE}
file_name <- system.file(file.path("extdata", "FORCIS_net_sample.csv"), package = "forcis")
net_data <- read.table(file_name, dec = ".", sep = ";")
```

```{r 'load-data-user', eval=FALSE}
# Import net data ----
net_data <- read_plankton_nets_data(path = "data")
```

**NB:** In this vignette, we use a subset of the plankton nets data, not the whole dataset.



## `select_taxonomy()`

The FORCIS database provides three different taxonomies: `LT` (lumped taxonomy), `VT` (validated taxonomy) and `OT` (original taxonomy). See the [associated data paper](https://doi.org/10.1038/s41597-023-02264-2) for further information.

After importing the data and before going any further, the next step involves choosing the taxonomic level for the analyses. Let's use the function `select_taxonomy()` to select the **VT** taxonomy (validated taxonomy):

```{r 'select-taxo'}
# Select taxonomy ----
net_data <- select_taxonomy(net_data, taxonomy = "VT")
```



## `select_columns()`

Because FORCIS data contains more than 100 columns, the function `select_columns()` can be used to lighten the `data.frame` to easily handle it and to speed up some computations. 

By default, only required columns listed in `get_required_columns()` and species columns will be kept.


```{r 'select-columns'}
# Select taxonomy ----
net_data <- select_columns(net_data)
```

But you can also use the argument `cols` to keep additional columns.



## `filter_by_month()`

The `filter_by_month()` function filters observations based on the **month of sampling**. It requires two arguments: the data and a numeric vector with values between 1 and 12.

```{r 'filter-by-month'}
# Filter data by sampling month ----
net_july_aug <- filter_by_month(net_data, months = 7:8)

# Number of original records ----
nrow(net_data)

# Number of filtered records ----
nrow(net_july_aug)
```


## `filter_by_year()`

The `filter_by_year()` function filters observations based on the **year of sampling**. It requires two arguments: the data and a numeric vector with the years of interest.

```{r 'filter-by-year'}
# Filter data by sampling year ----
net_90_20 <- filter_by_year(net_data, years = 1990:2020)

# Number of original records ----
nrow(net_data)

# Number of filtered records ----
nrow(net_90_20)
```


## `filter_by_bbox()`

The function `filter_by_bbox()` can be used to filter FORCIS data by a spatial bounding box (argument `bbox`).

Let's filter the plankton net data by a spatial rectangle located in the Indian ocean.

```{r 'filter-by-bbox'}
# Filter by spatial bounding box ----
net_data_bbox <- filter_by_bbox(net_data, bbox = c(45, -61, 82, -24))

# Number of original records ----
nrow(net_data)

# Number of filtered records ----
nrow(net_data_bbox)
```

Note that the argument `bbox` can be an object of class `bbox` (package `sf`) or a vector of four numeric values defining a square bounding box. If a vector of numeric values is provided, coordinates must be defined in the system WGS 84 (`epsg=4326`).



## `filter_by_ocean()`

The function `filter_by_ocean()` can be used to filter FORCIS data by one or several oceans (argument `ocean`).

Let's filter the plankton net data located in the Indian ocean.

```{r 'filter-by-ocean'}
# Filter by ocean name ----
net_data_indian <- filter_by_ocean(net_data, ocean = "Indian Ocean")

# Number of original records ----
nrow(net_data)

# Number of filtered records ----
nrow(net_data_indian)
```

Use the function `get_ocean_names()` to retrieve the name of World oceans according to the IHO Sea Areas dataset version 3 (used in this package).

```{r 'get-ocean-names'}
# Get ocean names ----
get_ocean_names()
```


## `filter_by_polygon()`

The function `filter_by_polygon()` can be used to filter FORCIS data a spatial polygon (argument `polygon`).

Let's filter the plankton net data by a spatial polygon defining boundaries of the Indian ocean.

```{r 'filter-by-polygon'}
# Import spatial polygon ----
file_name <- system.file(file.path("extdata", "IHO_Indian_ocean_polygon.gpkg"), 
                         package = "forcis")
indian_ocean <- sf::st_read(file_name, quiet = TRUE)

# Filter by polygon ----
net_data_poly <- filter_by_polygon(net_data, polygon = indian_ocean)

# Number of original records ----
nrow(net_data)

# Number of filtered records ----
nrow(net_data_poly)
```



## `filter_by_species()`

The function `filter_by_species()` allows to filter FORCIS data for one or more species.

It takes a `data.frame` and a vector of species names (argument `species`).

Let's subset plankton net data to only keep observations for *G. glutinata* and *C. nitida*.

```{r 'filter-by-species'}
# Filter by species ----
glutinata_nitida <- filter_by_species(net_data, 
                                      species = c("g_glutinata_VT", "c_nitida_VT"))

# Dimensions of original data ----
dim(net_data)

# Dimensions of filtered data ----
dim(glutinata_nitida)
```



## `reshape_data()`

The function `reshape_data()` converts FORCIS data into a long format.

```{r 'reshape-data'}
# Convert to long format ----
net_data_long <- reshape_data(net_data)

# Dimensions of original data ----
dim(net_data)

# Dimensions of reshaped data ----
dim(net_data_long)
```

Two columns have been created: `taxa` (taxon names) and `counts` (taxon counts).

```{r 'reshape-data-2'}
# Column names ----
colnames(net_data_long)
```
