---
title: "Select and filter data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Select and filter data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval     = FALSE,
  echo     = TRUE,
  comment  = "#>"
)
```

> **NOTE:** THIS IS A WORK IN PROGRESS.

The package `forcis` provides [a lot of functions](https://frbcesab.github.io/forcis/reference/index.html#select-and-filters-tools) to filter and select FORCIS data. This vignette shows how to use these functions.


## Setup

First, let's import the required packages.

```{r setup}
library(forcis)
```

Before going any further, we will download the latest version of the FORCIS database.

```{r 'download-db', eval=FALSE}
# Create a data/ folder ----
dir.create("data")

# Download latest version of the database ----
download_forcis_db(path = "data", version = NULL)
```

The vignette will use the PUMP data of the FORCIS database. Let's import the latest release of the data.

```{r 'load-data', echo=FALSE}
file_name <- system.file(file.path("extdata", "FORCIS_pump_sample.csv"), package = "forcis")
pump_data <- vroom::vroom(file_name, delim = ";", show_col_types = FALSE)
```

```{r 'load-data-user', eval=FALSE}
# Import pump data ----
pump_data <- read_pump_data(path = "data")
```

**NB:** In this vignette, we use a subset of the PUMP data, not the whole dataset.



## `select_columns()`

Add an illustration of `select_columns()`

[...]



## `select_taxonomy()`

Add an illustration of `select_taxonomy()`

[...]



## `filter_by_species()`

Add an illustration of `filter_by_species()`

[...]



## `filter_by_month()`

Add an illustration of `filter_by_month()`

[...]



## `filter_by_year()`

Add an illustration of `filter_by_year()`

[...]



## `filter_by_bbox()`

Add an illustration of `filter_by_bbox()`

[...]



## `filter_by_ocean()`

Add an illustration of `filter_by_ocean()`

[...]



## `filter_by_polygon()`

Add an illustration of `filter_by_polygon()`

[...]
